Altering Backward Pass Gradients to improve Convergence

### 2022-11-17 0100.tex
Mod2

\documentclass[runningheads]{llncs}
\usepackage[T1]{fontenc}
\usepackage{framed}
\usepackage{multirow}
\usepackage{amssymb}
\usepackage{latexsym}
\usepackage{url}
\usepackage{xcolor}
\usepackage{subcaption}
\usepackage{amsmath}
\usepackage{listings}
\usepackage{booktabs}
\usepackage{pifont}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{color}
\graphicspath{{images/}}
\renewcommand\UrlFont{\color{blue}\rmfamily}
\newcommand{\norm}[1]{\left\lVert#1\right\rVert}
\NewDocumentCommand{\codeword}{v}{\texttt{\textcolor{orange}{#1}}}
\newcommand{\cmark}{\ding{51}} \newcommand{\xmark}{\ding{55}}
\definecolor{newcolor}{rgb}{.8,.349,.1}
\definecolor{codegreen}{rgb}{0.7,0.7,0.7}
\definecolor{codegray}{rgb}{0.5,0.5,0.5}
\definecolor{codepurple}{rgb}{0.58,0,0.82}
\definecolor{backcolour}{rgb}{1,1,1}
\lstdefinestyle{mystyle}{ backgroundcolor=\color{backcolour},   
commentstyle=\color{codegreen}, keywordstyle=\color{magenta},
numberstyle=\tiny\color{codegray}, stringstyle=\color{codepurple},
basicstyle=\ttfamily\footnotesize, breakatwhitespace=false,         
breaklines=true,                 
captionpos=b,                    
keepspaces=true,                 
numbersep=5pt,                  
showspaces=false,                
showstringspaces=false, showtabs=false,                  
tabsize=2 }
\lstset{style=mystyle}

\begin{document}
\title{Altering Backward Pass Gradients to improve Convergence}
\author{Bishshoy Das\inst{1}\orcidID{0000-0002-6525-4015}\and Milton Mondal\inst{1}\orcidID{xxxx-xxxx-xxxx-xxxx}\and Brejesh Lall\inst{3}\orcidID{xxxx-xxxx-xxxx-xxxx}\and Shiv Dutt Joshi\inst{3}\orcidID{0000-0002-0326-0524}\and Sumantra Dutta Roy\inst{3}\orcidID{0000-0002-2141-5067}}
\authorrunning{B. Das et al.}
\institute{Indian Institute of Technology Delhi, New Delhi - 110016, India \\ \email{\{bishshoy.das, milton.mondal, brejesh, sdjoshi, sumantra\}@ee.iitd.ac.in}}
\maketitle

@@

Backpropagation is traditionally used to train deep neural networks
\cite{lillicrap2020backpropagation}. Gradients are computed using basic calculus
principles to adjust the weights during backpropagation \cite{lecun1988theoretical}.
Alternatives to traditional gradients has rarely been studied in the literature
hitherto. In normal training procedures, gradients are computed immediately in the
backward pass utilizing the values obtained in the forward pass. This makes the backward
pass coupled with the forward pass. However, decoupling the backward pass from the
forward pass by modifying the gradients to improve training efficiency and final
convergent accuracy has hardly been explored. In this paper we explore the landscape of
decoupling the forward pass from the backward pass by altering the gradients and
subsequently updating the network's parameters with the modified gradients. There are
several ways to achieve gradient modification in the backward pass. We discuss a few
techniques in Fig. \ref{fig:gradient_altering}. One could modify the gradients at either
\textbf{(I)} multiple points in the backward computation graph or \textbf{(II)} at a
point very early in the backward graph and allow this changed gradient to also affect
the rest of the network's gradients as it backpropagates through the backward
computation graph.

\begin{figure}[t]
\centering
\includegraphics[width=\textwidth]{PGT_Types}
\caption{ Different ways of altering gradients in the backward pass. Blue blocks denote
a CNN's different layers (convolutional, linear, batchnorm, softmax, etc.). Orange
blocks indicate the backward graph with unmodified gradients. Green blocks represent
transformation functions, while red blocks indicate transformed gradients. (a)
\textbf{Type 0:} Backpropagation without any gradient modification. (b) \textbf{Type I:}
Altering gradients after they are all computed. (c) \textbf{Type II:} Altering gradients
at an earlier stage in the backward computation graph, subsequently leading to
alteration of all other gradients. }
\label{fig:gradient_altering}
\end{figure}

@@

\textbf{Type I:} Gradient modification at multiple points independently: In this method,
the gradients are first computed using standard procedure and then individually altered
as shown in Fig. \ref{fig:gradient_altering}(b). Gradient clipping
\cite{pascanu2013difficulty} and Adaptive gradient clipping \cite{brock2021high} are
examples of such modifications. In both of these methods, the gradients are first
computed using standard rules and then they are modified using some function. It can be
described as:

@@

Type I modification is computationally more expensive than Type II modification as it
requires altering the gradients of each and every parameter individually. Type II
modification recomputes gradients at each and every location through the natural flow of
backpropagation. We propose PowerGrad Transform (PGT) which is a type II gradient
modification method. With virtually no additional compute or memory requirement,
PowerGrad Transform enhances classification performance of a convolutional neural
network, by leading the network to arrive at a higher quality optima at convergence, at
which both training and test accuracies improve.

PowerGrad Transform modifies the gradients at the softmax layer, which is the earliest
part of most convolutional neural network's backward computation graph. By changing the
gradient at this stage, all other gradients for all other parameters in the network such
as the linear layer, batchnorm statistics and convolutional layer's filter parameters
are affected. We analyze and explore the behaviour of such alteration both
mathematically and through empirical experiments conducted on networks without
batch-normalization. Experimentally, we see that PGT improved training and test
accuracies for both normalized and non-normalized networks. PGT improves a network's
learning capacity and results in a better fit of the training data.

@@

\item We introduce PowerGrad Transform, which decouples the backward and the forward
passes of neural network training and enables a considerably better fit to the dataset,
as assessed by both training and test accuracy measures. PGT is a performance
enhancement method that alters the gradients in the backward pass before the update step
leading to accelerated training and a significant boost in the network's predictive
performance.

\item We perform theoretical analysis of the properties of the PowerGrad transformation
(section \ref{sec:pgt_prop}) and explore its effect on weight parameters and gradients
(section \ref{sec:pgt_weight_gradients}), logits (section
\ref{sec:pgt_logit_derivative}), (section \ref{sec:Effe}), loss values (section
\ref{sec:Effe}) and class separation (section \ref{sec:pgt_class_separation}).

\item We study experimentally the degenerate behaviour of non-BN networks that is often
seen during the normal training procedure, particularly at larger batch sizes. With PGT,
we see improved gradient behaviour and a decreased likelihood of the weights attaining
degenerate states.

\item We provide complete results from a variety of models (non-BN and BN ResNets,
SE-ResNets) using the ImageNet dataset. It helps the network to improve its learning
capabilities by locating a more optimum convergence point. Additionally, we conduct an
ablation study and compare its impacts to those of regularization approaches such as
label smoothing.

@@

We examine techniques that are related to gradient modification and emphasize the
distinction between them and our proposed method.

Gradient Clipping (GC): Gradient Clipping \cite{pascanu2013difficulty}, often used in
natural language processing methods \cite{merity2017regularizing}, is a technique that
involves changing or clipping the gradients with respect to a predefined threshold value
during backward propagation through the network and updating the weights using the
clipped gradients \cite{zhang2019gradient} \cite{smith2020generalization}. By rescaling the
gradients, the weight updates are likewise rescaled, significantly reducing the risk of
an overflow or underflow \cite{pascanu2012understanding}. GC can be used for training
networks without batch-normalization.

The formulation is as follows: If $G$ is a gradient vector which is $\partial L /
\partial \theta$ (the gradient of the loss $L$ with respect to the parameters $\theta$).
In clip by value, the gradients are modified as:

\begin{equation}
G \rightarrow
\begin{cases}
\lambda \frac{G}{\norm{G}} & \text{if } \norm{G} > \lambda, \\
G & \text{otherwise}
\end{cases}
\end{equation}

At larger batch sizes, the clipping threshold in GC becomes highly sensitive and
requires extensive finetuning for various models, batch sizes, and learning rates. As we
demonstrate later in our studies, GC is not as effective in improving the performance of
non-normalized networks. AGC performs better than GC in non-normalized networks.
However, we show that PGT outperforms both in such networks.

Adaptive Gradient Clipping (AGC): Adaptive Gradient Clipping \cite{brock2021high} is
developed to further enhance backward pass gradients than what is performed by GC. It
takes into account the fact that the ratio of the gradient norm to the weight norm can
provide an indication of the expected change in a single step of optimization. Here the
normalized gradient is clipped by $\lambda$, which means that large weights can have
large gradients and still the normalized gradient can be within $\lambda$. If it is not
within the threshold, then their gradient is normalized by the ratio of norm of the
gradient and norm of the weight in order to avoid gradient explosion. If $W^l$ is the
weight matrix of the $l^\textit{th}$ layer, and $G^l$ is its gradient in the backward
pass, then the following equation is used to modify the gradient of the $i^\textit{th}$
filter before its update:

\begin{equation}
G_{i}^{l}\rightarrow
\begin{cases}
\lambda \frac{\norm{W_{i}^{l} }_{F}^{*}}{\norm{G_{i}^{l} }_{F}} G_{i}^{l} & if\ \frac{\norm{G_{i}^{l} }_{F}}{\norm{W_{i}^{l} }_{F}^{*}} \  >\ \lambda \\
G_{i}^{l} & otherwise
\end{cases}
\end{equation}

The hyperparameter $\lambda$ of AGC is less sensitive to changes in batch size and depth
as observed in \cite{brock2021high}. However, when applied to vanilla residual networks
(ResNet-18), the performance is still afar from normalized variants as demonstrated by
our experiments. In most cases, AGC is used for training networks without
batch-normalization. Our experiments demonstrate that PGT outperforms AGC, when used
independently. When used together, it can further improve a network's performance
(section \ref{sec:Empi}).

Label Smoothing: Label smoothing, introduced by Szegedy et al.
\cite{szegedy2016rethinking}, utilizes smoothing of the ground truth labels as a method
to impose regularization on the logits and the weights of the neural network.

The formulation is as follows. If $q_i$ is the value at the $i^{th}$ index of the one
hot encoded vector of the ground truth label, then the transformed distribution of the
labels $q_i'$ is: \begin{equation} q_i' = \begin{cases} 1-\epsilon & \text{if } i = y,
\\ \frac{\epsilon}{K-1} & otherwise \end{cases} \end{equation} where $\epsilon$ is a
hyperparameter with a value in $[0,1]$ and $y$ denotes the index of the correct class.

Müller et al. investigates how label smoothing work in \cite{DBLP:conf/nips/MullerKH19}.
Using visualizations of penultimate layer activations, they demonstrate that label
smoothing calibrates the networks predictions and aligns them with the accuracies of the
predictions. Parallel works in label smoothing include Pairwise Confusion
\cite{dubey2018pairwise}, combatting Label Noise \cite{reed2014training}, achieving
regularization through intentional label modification \cite{xie2016disturblabel}.

PowerGrad Transform works in a much different way. PGT does not smooth the ground truth
labels, rather it modifies the gradient by smoothing the predicted probability vector.
As we show in section \ref{sec:Math}, our proposed transformation has the opposite
effect of label smoothing. From our studies, we show that PGT in fact leads to larger
generalization gap between the training and test dataset, although it does lead to an
increase in both training and test accuracies at convergence. Coupled with PGT, label
smoothing can narrow the generalization gap, while retaining the benefits of PGT.

Knowledge Distillation: The formulation of PowerGrad Transform is akin to temperature
based logit suppression of \cite{hinton2015distilling}. Knowledge distillation
\cite{DBLP:conf/nips/BaC14} is a process in which two networks are trained. First a
teacher network is trained on a given dataset and then the soft-labels (the predicted
probabilities) of the teacher network are used to train another network, called the
student network. In similar style of label smoothing, knowledge distillation aims to
generate smooth gradient behaviour by forcing the network not to overpredict on any
single image. As in case of label smoothing, the student network's weights are
automatically penalized if the network assigns a probability value higher than the soft
labels generated by the teacher network. Variants of knowledge distillation include
self-distillation \cite{zhang2019your} \cite{wang2021memory}, identical student network
distillation \cite{furlanello2018born}, channel distillation \cite{ge2019distilling},
regularizing wrong predictions \cite{yun2020regularizing} and representation or
embedding based knowledge distillation \cite{aguilar2020knowledge} \cite{yao2018deep}
\cite{passalis2018unsupervised}. Distillation is applied to various tasks such as model
compression \cite{wang2020real}, face recognition \cite{ge2018low}, enhancing ensemble
performance \cite{zhang2018deep}, interpreting neural networks \cite{liu2018improving}.
Efficacy of distillation is extensivly studied in \cite{cho2019efficacy}
\cite{yuan2020revisiting}. Applications of logit suppression include: deep metric
learning \cite{zhai2018classification}, non-parametric neighborhood component analysis
\cite{wu2018improving}, sequence-to-sequence modelling \cite{chorowski2016towards},
reinforcement learning \cite{he2018determining} and face recognition
\cite{liu2017sphereface} \cite{wang2018cosface}.

The key difference between PGT and distillation methods is that in the latter, the
transformation is applied in the forward pass, while PGT is a backward pass modification
only. Also, in distillation settings the temperature parameter is a part of the
network's computation graph. In the case of PowerGrad Transform, we directly tamper the
gradients without introducing any change in the forward pass. It is a more generalized
way of arriving at altered gradients in the backward pass. Moreover we show that
modifying the gradients in the backward pass can improve performance in many cases. The
rise in training accuracy indicates that PGT enables the network to learn more from the
training data.

Self-knowledge distillation is used in many papers in different ways, which mainly
focuses on improving the performance of the model by either creating multiple ensemble
student models \cite{zhang2018deep}, using different distorted versions of same data
with two different versions of a single network which guides one another
\cite{xu2019data} or injecting multiple sub-modules in the network to improve its
capacity \cite{zhang2019your}. However none of these methods use normal traditional
training without introducing sub-modules in the architecture. PGT differs from
self-knowledge distillation as it neither introduces any additional sub-modules nor
creates different ensembles to improve the performance of the model. PGT follows the
standard neural network training mechanism with a single change which is altering the
last layer's gradients in the backward pass.

Other techniques of weight update include slot machines \cite{aladago2021slot}, where
random weights are generated and carefully selected during backpropagation in a way that
minimizes the loss function.

@@

\begin{figure}[t]
\centering
\begin{subfigure}{.5\textwidth}
\centering
\includegraphics[width=0.98\textwidth]{prob_transform_unordered}
\caption{Unordered distribution}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
\centering
\includegraphics[width=0.98\textwidth]{prob_transform_ordered}
\caption{Rank ordered distribution}
\end{subfigure}
\caption{ Illustration of the impact of PowerGrad Transform on probability values of an
arbitrary distribution (number of classes, $C=10$). $\alpha=1$ indicates the unmodified
distribution. As $\alpha$ reduces from $1$ to $0$, the probability distributions (left:
unordered or right: ordered) becomes flatter and flatter, approaching the uniform
distribution of $1/C$ for every index. }
\label{fig:prob_plots}
\end{figure}

The PowerGrad Transform technique is described in this section. It is a general
technique and can be applied to any neural network. A neural network with parameters $W$
generates $C$ logits denoted by $z$ for every input vector $x$. $z$ is given as $z=Wx$.
Then a set of probability values $p_i$ are generated from the logits using a softmax
function which is defined as $p_i=\frac{e^{z_i}}{\sum _{j=1}^C e^{z_j}}$. $p_i$ and
$z_i$ represent the predicted probability values and the logits for the $i^{th}$ class
respectively. Following this step, the loss function is invoked and the loss between the
predicted probability values the ground truth labels (which is also a probability
distribution) is calculated. If the loss function is cross-entropy loss, then the value
of the loss is given as $L=-\sum _{i=1}^C q_i \log \left(p_i\right)$ where $q_i$ is the
ground truth label of the $i^{th}$ class for a particular training example. By standard
gradient update rule, we can calculate the gradient of the loss with respect to the
logits which takes the form $\frac{\partial L}{\partial z_i}=p_i-q_i$.

@@

\begin{equation} \hspace*{3.5cm} p_i'=\frac{p_i^{\alpha }}{\sum _{j=1}^C
p_j^\alpha} \ \ \ \ \ \ \ \ i=1,\dots,C\ \ \ \ \ 0 \leq \alpha\leq 1
\label{transformed_probabilities} \end{equation}

@@

\begin{equation} \widehat{\frac{\partial L}{\partial z_i}}=p_i'-q_i
\label{PGT_logit_derivative}
\end{equation}

@@

A code snippet of the algorithm implemented using PyTorch \cite{NEURIPS2019_9015} is
given in the appendix. In PyTorch, the available method to modify gradients in the
backward pass is using the \codeword{register_hook} function which is called in the
\codeword{forward} function. We explore the effect of this change from a theoretical
standpoint in section \ref{sec:Math}. In section \ref{sec:Expe}, we offer empirical data
and experiments to support the proposed method.

\section{Analysis of the PowerGrad transformation}
\label{sec:Math}

In this section, we first describe the properties of the  PowerGrad Transform (PGT) and
then we highlight how these suitable properties of PGT helps the neural network to
improve the performance. We use the same setup as described in section \ref{sec:Powe}.
To explore the properties of PGT, we start by investigating the effect of the transform
on the softmax probabilities.

\subsection{Properties}
\label{sec:pgt_prop}

@@

We call this threshold, the \textit{stationary threshold}. The stationary threshold is
that value of $p_i$ that does not change after the transformation. Therefore, when $p_i$
is greater than the \textit{stationary threshold}, $p_i' < p_i$.

@@

We conclude from the analysis that the stationary threshold is a monotonic
non-decreasing function with respect to $\alpha$. Also the derivative of PGT function
with respect to the true probabilities is non-negative which in turn means that the
transformation is an order-preserving map. All values greater than the threshold move
towards the threshold after transformation and all values below the threshold also move
towards the threshold, and the threshold itself moves monotonically towards $1/C$ as
$\alpha$ is decreased from $1$ to $0$. This concludes that the transformation smooths
the original distribution.

\subsection{Effects of PGT on logit, weight gradients and class separation}

@@

\subsection{Experiments on convolutional neural networks (with batch-normalization)}
\label{sec:bn}

\begin{figure}[!t]
\centering
\begin{subfigure}{.33\textwidth}
\centering
\includegraphics[width=0.98\textwidth]{acc_vs_epoch_r18}
\caption{Plot of training and test accuracies (ResNet-18)}
\end{subfigure}%
\begin{subfigure}{.33\textwidth}
\centering
\includegraphics[width=0.98\textwidth]{acc_vs_epoch_r50}
\caption{Plot of training and test accuracies (ResNet-50)}
\end{subfigure}%
\begin{subfigure}{.33\textwidth}
\centering
\includegraphics[width=0.98\textwidth]{acc_vs_epoch_r50_60epochs}
\caption{Plot of test accuracies (ResNet-50)}
\end{subfigure}
\caption{ Log-linear plots of training and test accuracies and comparison with baseline
of batch-normalized variants: \textbf{(a)} ResNet-18 ($\alpha=0.25$), \textbf{(b)}
ResNet-50 ($\alpha=0.3$). For better visualization of accuracies obtained in the final
iterations, we plot both axes in log scale. PowerGrad Transform improves upon the
baseline in all cases. The corresponding accuracies are mentioned in Table
\ref{tab:imagenet_table}. \textbf{(c)} Training speed comparison between PGT ($60$
epochs) and baseline ($100$ epochs). They both converge to the same test accuracy
($76.5\%$) on ImageNet-1K. PGT's accelerated training saves $40\%$ of the epoch budget.
}
\label{fig:metrics}
\end{figure}

>>>>

\documentclass[runningheads]{llncs}
\usepackage[T1]{fontenc}
\usepackage{framed}
\usepackage{multirow}
\usepackage{amssymb}
\usepackage{latexsym}
\usepackage{url}
\usepackage{xcolor}
\usepackage{subcaption}
\usepackage{amsmath}
\usepackage{listings}
\usepackage{booktabs}
\usepackage{pifont}
\usepackage{wrapfig}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{color}
\graphicspath{{images/}}
\raggedbottom
\renewcommand\UrlFont{\color{blue}\rmfamily}
\newcommand{\norm}[1]{\left\lVert#1\right\rVert}
\NewDocumentCommand{\codeword}{v}{\texttt{\textcolor{orange}{#1}}}
\newcommand{\cmark}{\ding{51}} \newcommand{\xmark}{\ding{55}}
\definecolor{newcolor}{rgb}{.8,.349,.1}
\definecolor{codegreen}{rgb}{0.7,0.7,0.7}
\definecolor{codegray}{rgb}{0.5,0.5,0.5}
\definecolor{codepurple}{rgb}{0.58,0,0.82}
\definecolor{backcolour}{rgb}{1,1,1}
\lstdefinestyle{mystyle}{ backgroundcolor=\color{backcolour},   
commentstyle=\color{codegreen}, keywordstyle=\color{magenta},
numberstyle=\tiny\color{codegray}, stringstyle=\color{codepurple},
basicstyle=\ttfamily\footnotesize, breakatwhitespace=false,         
breaklines=true,                 
captionpos=b,                    
keepspaces=true,                 
numbersep=5pt,                  
showspaces=false,                
showstringspaces=false, showtabs=false,                  
tabsize=2 }
\lstset{style=mystyle}

\begin{document}
\title{Altering Backward Pass Gradients to improve Convergence}
\author{Bishshoy Das\inst{1}\orcidID{0000-0002-6525-4015}\and Milton Mondal\inst{1}\orcidID{xxxx-xxxx-xxxx-xxxx}\and Brejesh Lall\inst{3}\orcidID{xxxx-xxxx-xxxx-xxxx}\and Shiv Dutt Joshi\inst{3}\orcidID{0000-0002-0326-0524}\and Sumantra Dutta Roy\inst{3}\orcidID{0000-0002-2141-5067}}
\authorrunning{Das et al.}
\institute{Indian Institute of Technology Delhi, New Delhi - 110016, India \\ \email{\{bishshoy.das, milton.mondal, brejesh, sdjoshi, sumantra\}@ee.iitd.ac.in}}
\maketitle

@@

Backpropagation is traditionally used for training deep neural networks
\cite{lillicrap2020backpropagation}. Gradients are computed using basic calculus
principles to adjust the weights during backpropagation \cite{lecun1988theoretical}.
Alternatives to traditional gradients has rarely been studied in the literature
hitherto. In normal training procedures, gradients are computed immediately in the
backward pass utilizing the values obtained in the forward pass. This makes the backward
pass coupled with the forward pass. However, decoupling the backward pass from the
forward pass by modifying the gradients to improve training efficiency and final
convergent accuracy has hardly been explored. In this paper we explore the landscape of
decoupling the forward pass from the backward pass by altering the gradients and
subsequently updating the network's parameters with the modified gradients. There are
several ways to achieve gradient modification in the backward pass. We discuss a few
techniques in Fig. \ref{fig:gradient_altering}.

\begin{figure}[t]
\centering
\includegraphics[width=\textwidth]{PGT_Types}
\caption{ Different ways of altering gradients in the backward pass. Blue blocks denote
a different layers. Orange blocks indicate the backward graph with unmodified gradients.
Green blocks represent transformation functions, while red blocks indicate transformed
gradients. }
\label{fig:gradient_altering}
\vspace{-0.5cm}
\end{figure}

@@

\textbf{Type I:} Independent gradient modification at multiple points: In this method,
the gradients are first computed using standard procedure and then individually altered
as shown in Fig. \ref{fig:gradient_altering}(b). Gradient clipping
\cite{pascanu2013difficulty} and Adaptive gradient clipping \cite{brock2021high} are
examples of such modifications. In both of these methods, the gradients are first
computed using standard rules and then they are modified using some function. It can be
described as:

@@

Type I is computationally more expensive than type II as it requires altering the
gradients of each and every parameter individually. Type II modification recomputes
gradients at each and every location through the natural flow of backpropagation. We
propose PowerGrad Transform (PGT), a type II method that modifies the gradients at the
softmax layer.

@@

\item We propose PowerGrad Transform, which decouples the backward and the forward
passes of neural network training and enables a considerably better fit to the dataset.
PGT is a performance enhancement method that alters the gradients in the backward pass
before the update step leading to accelerated training and a significant boost in the
network's predictive performance.

\item We perform theoretical analysis of the properties of the PowerGrad transformation
(section \ref{sec:pgt_prop}) and explore its effect on weight parameters and gradients,
logits, loss values and class separation (section \ref{sec:pgt_effects}).

\item We provide complete results from a variety of models (non-BN and BN ResNets,
SE-ResNets) using the ImageNet dataset. We empirically conclude that PGT helps a network
to improve by locating a more optimum convergence point. Additionally, we conduct
ablation studies and compare its effects to those of regularization methods.

@@

Gradient Clipping \cite{pascanu2013difficulty} is a technique that involves changing or
clipping the gradients with respect to a predefined threshold value during backward
propagation through the network and updating the weights using the clipped gradients
\cite{zhang2019gradient}\cite{smith2020generalization}. By rescaling the gradients, the
weight updates are likewise rescaled, significantly reducing the risk of an overflow or
underflow \cite{pascanu2012understanding}. GC can be used for training networks without
batch-normalization. At larger batch sizes, the clipping threshold in GC becomes highly
sensitive and requires extensive finetuning for various models, batch sizes, and
learning rates. As we demonstrate later in our studies, GC is not as effective in
improving the performance of non-normalized networks. Adaptive Gradient Clipping
\cite{brock2021high} is developed to further enhance backward pass gradients than what
is performed by GC. It takes into account the fact that the ratio of the gradient norm
to the weight norm can provide an indication of the expected change in a single step of
optimization. Label smoothing, introduced by Szegedy et al.
\cite{szegedy2016rethinking}, utilizes smoothing of the ground truth labels as a method
to impose regularization on the logits and the weights of the neural network. AGC
performs better than GC in non-normalized networks. However, we show that PGT
outperforms both in such networks.

Knowledge distillation \cite{hinton2015distilling} is a process in which two networks
are trained with hard and soft labels alternatively. Variants of knowledge distillation
include self-distillation \cite{zhang2018deep}, identical student network distillation
\cite{furlanello2018born}, channel distillation \cite{ge2019distilling}, regularizing
wrong predictions \cite{yun2020regularizing} and representation or embedding based
knowledge distillation \cite{yao2018deep}\cite{aguilar2020knowledge}. The key difference between PGT and knowledge
distillation is that in the latter, the transformation is applied in the forward pass,
while PGT is a backward pass modification only. Also, in distillation settings the
temperature parameter is a part of the network's computation graph. In the case of
PowerGrad Transform, we directly tamper the gradients without introducing any change in
the forward pass. PGT differs from self-knowledge distillation as it neither introduces
any additional sub-modules nor creates different ensembles to improve the performance of
the model. PGT follows the standard neural network training mechanism with modified
gradients.

@@

\begin{figure}[t]
\centering
\begin{subfigure}{.5\textwidth}
\centering
\includegraphics[width=0.98\textwidth]{prob_transform_unordered}
\caption{Unordered distribution}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
\centering
\includegraphics[width=0.98\textwidth]{prob_transform_ordered}
\caption{Rank ordered distribution}
\end{subfigure}
\caption{ Illustration of the impact of PowerGrad Transform on probability values of an
arbitrary distribution (number of classes, $C=10$). $\alpha=1$ indicates the unmodified
distribution. As $\alpha$ reduces from $1$ to $0$, the probability distributions (left:
unordered or right: ordered) becomes flatter and flatter, approaching the uniform
distribution of $1/C$ for every index. }
\label{fig:prob_plots}
\vspace{-0.5cm}
\end{figure}

A neural network with parameters $W$ generates $C$ logits denoted by $z$ for every input
vector $x$. $z$ is given as $z=Wx$. Then a set of probability values $p_i$ are generated
from the logits using a softmax function which is defined as $p_i=\frac{e^{z_i}}{\sum
_{j=1}^C e^{z_j}}$. $p_i$ and $z_i$ represent the predicted probability values and the
logits for the $i^{th}$ class respectively. Following this step, the loss function is
invoked and the loss between the predicted probability values the ground truth labels
(which is also a probability distribution) is calculated. If the loss function is
cross-entropy loss, then the value of the loss is given as $L=-\sum _{i=1}^C q_i \log
\left(p_i\right)$ where $q_i$ is the ground truth label of the $i^{th}$ class for a
particular training example. By standard gradient update rule, we can calculate the
gradient of the loss with respect to the logits which takes the form $\frac{\partial
L}{\partial z_i}=p_i-q_i$.

@@

\vspace{-0.5cm}
\begin{equation} \hspace*{3.5cm} p_i'=\frac{p_i^{\alpha }}{\sum _{j=1}^C
p_j^\alpha} \ \ \ \ \ \ \ \ i=1,\dots,C\ \ \ \ \ 0 \leq \alpha\leq 1
\label{transformed_probabilities} \end{equation}

@@

\vspace{-0.5cm}
\begin{equation} \widehat{\frac{\partial L}{\partial z_i}}=p_i'-q_i
\label{PGT_logit_derivative}
\end{equation}

@@

\subsection{Properties of the PowerGrad transformation}
\label{sec:pgt_prop}

We use the same setup as described in section \ref{sec:Powe}. To explore the properties
of PGT, we start by investigating the effect of the transform on the softmax
probabilities.

@@

\vspace{-0.5cm}
\begin{equation} \begin{split} p_i' \geq p_i & \text{,\ \ if } p_i \leq
\Big(\sum _{j=1}^C p_j^{\alpha }\Big)^{\frac{1}{\alpha-1}} \\ \hspace*{1cm}
p_i' < p_i & \text{,\ \ if } p_i > \Big(\sum _{j=1}^C
p_j^{\alpha}\Big)^{\frac{1}{\alpha-1}} \end{split} \label{eqn:threshold}
\end{equation}

\vspace{-0.2cm}
We call this threshold, the \textit{stationary threshold}. The stationary threshold is
that value of $p_i$ that does not change after the transformation. Therefore, when $p_i$
is greater than the \textit{stationary threshold}, $p_i' < p_i$.

@@

We provide the proofs in the supplementary section and conclude that the stationary
threshold is a monotonic non-decreasing function with respect to $\alpha$. Also the
derivative of PGT function with respect to the true probabilities is non-negative which
in turn means that the transformation is an order-preserving map. All values greater
than the threshold move towards the threshold after transformation and all values below
the threshold also move towards the threshold, and the threshold itself moves
monotonically towards $1/C$ as $\alpha$ is decreased from $1$ to $0$. This concludes
that the transformation smooths the original distribution.

\subsection{Effects of PGT on logit, weight gradients and class separation}
\label{sec:pgt_effects}

@@

\begin{figure}[!t]
\centering
\begin{subfigure}{.33\textwidth}
\centering
\includegraphics[width=0.98\textwidth]{acc_vs_epoch_r18}
\caption{Plot of training and test accuracies (ResNet-18)}
\end{subfigure}%
\begin{subfigure}{.33\textwidth}
\centering
\includegraphics[width=0.98\textwidth]{acc_vs_epoch_r50}
\caption{Plot of training and test accuracies (ResNet-50)}
\end{subfigure}%
\begin{subfigure}{.33\textwidth}
\centering
\includegraphics[width=0.98\textwidth]{acc_vs_epoch_r50_60epochs}
\caption{Plot of test accuracies (ResNet-50)}
\end{subfigure}
\caption{ Log-log plots of training and test accuracies and comparison with baseline of
batch-normalized variants: \textbf{(a)} ResNet-18 ($\alpha=0.25$), \textbf{(b)}
ResNet-50 ($\alpha=0.3$). \textbf{(c)} Training speed comparison between PGT ($60$
epochs) and baseline ($100$ epochs). They both converge to the same test accuracy
($76.5\%$) on ImageNet-1K. PGT's accelerated training saves $40\%$ of the epoch budget.
}
\label{fig:metrics}
\vspace{-0.5cm}
\end{figure}

We perform experiments on different variants ResNets using the ImageNet-1K dataset
\cite{deng2009imagenet}. All models are trained on four V100 GPUs with a batch size of
$1024$. We utilize a common set of hyperparameters for all experiments, which are as
follows: 100 epoch budget, 5 epochs linear warmup phase beginning with a learning rate
of $4\times 10^{-4}$ and ending with a peak learning rate of $0.4$, a momentum of $0.9$
and weight decay of $5\times 10^{-4}$, the SGD Nesterov optimizer and mixed precision.
In our studies, we employ either a step scheduler (dividing the learning rate by $10$ at
the $30^{th}$, $60^{th}$, and $90^{th}$ epochs) or a cosine decay scheduler
\cite{loshchilov2016sgdr}. We find $\alpha=0.25$ and $\alpha=0.05$ to be good choices
for ResNet-18 and ResNet-50, though larger values such as $\alpha=0.3$ also have good
performance as well. The experimental results are shown in Table
\ref{tab:imagenet_table}, and we mention the value of the PGT hyperparmeter ($\alpha$)
in each experiment. We explore different values of $\alpha$ in section \ref{sec:Abla}
and provide detailed grid plots in the supplementary section\footnote{Reproducible code,
training recipes for the experiments, pretrained checkpoints and training logs are
provided at: \url{ https://github.com/skalien/power-grad-transform }.}.
